Parallel Adaptive Importance Sampling

### 2017-01-16 old_intro.tex
New intro
New intro written.

@@

In this example we compare the naively parallelised RWMH algorithm with its
PAIS variant, the PAIS-RW algorithm. The PAIS algorithm is implemented using the
ETPF to perform the resampling step. We assess the performance of the PAIS algorithm
using the relative $L^2$ error defined in \eqref{eqn:L2_error}, as well as the
relative error in the first moment \eqref{eq:34567}.

Since we are comparing against naively parallelised MH algorithms,
we also need to decide which statistics $T(\beta)$ provide the best criterions
for obtaining the optimal scaling parameters. In the examples which follow, we have
optimised the naively parallelised RWMH algorithm using the optimal
acceptance rate $\hat{\alpha} = 0.5$. This value differs from the theoretical
asymptotic value of 0.234 which applies in higher dimensions, but this higher acceptance rate is commonly used for one dimensional Gaussian posteriors~\cite{rosenthal2011optimal}. To find the optimal
scaling parameter we minimise the statistic
\[
	T_{\text{MH}}(\beta) = \left| \frac{N_{\text{acc}}(\beta)}{N_{\text{total}}} - \hat{\alpha} \right|,
\]
where $N_{\text{acc}}(\beta)$ is the number of accepted moves and
$N_{\text{total}}$ is the total number of samples produced. For the
PAIS algorithm, we maximise the effective sample size as
discussed in Section~\ref{sec:ess}.

@@

Consider the simple case of a linear observation operator $\G(x) = x$, where the prior on $x$ and the observational noise follow Gaussian distributions. Then, following \eqref{eqn:like}, the Gaussian posterior has the form
\begin{equation}\label{eqn:Gaussian posterior}
	\text{law}(\mu_D) = \pi(x|D) \propto \exp\left(-\frac{1}{2}\big\|x - D\big\|^2_{\sigma^2} - \frac{1}{2}\big\|x\big\|^2_{\tau^2}\right),
\end{equation}
where $\sigma^2$ and $\tau^2$ are the variances of the
observational noise and prior distributions respectively. In the
numerics which follow, we choose $\tau^2 =0.01$ and $\sigma^2 = 0.01$, and
we observe $x_\text{ref}=4$ noisily such that
\[
D = \mathcal{G}(x) + \eta \sim \mathcal{N}(\mathcal{G}(x_\text{ref}),\sigma^2) = \mathcal{N}(4,0.01).
\]
These values result in a posterior density in which the vast majority
of the density is out in the tails of the
prior distribution. The Kullback-Leibler (KL)
divergence, which gives us a measure of how different the prior and
posterior are, is $D_{KL}(\mu_D || \mu_0) = 4.67$ for this
problem. A KL divergence of zero indicates that two distributions are
identical almost everywhere.

@@

\begin{figure}[htb]
\centering
\subfigure[RWMH algorithm.]{\includegraphics[width=0.45\textwidth]{"figures/G2_rwmho"}}
\subfigure[PAIS-RW algorithm.]{\includegraphics[width=0.45\textwidth]{"figures/G2_paiso"}}
\caption{Finding optimal values of $\beta$ for the problem in Section~\ref{sec:problem 1}. The setup is as in Section~\ref{sec:Implementation P1}. Resampling is performed using the ETPF.}
\label{fig:P1 opt beta}
\end{figure}

@@

Figure~\ref{fig:P1 opt beta} (b) shows the effective sample size ratio
compared to the error analysis and the variance of the weights. The
relative $L^2$ error graph is noisy, but it is clear that the maximum
in the effective sample size and the minimum in the variance of the
weights are both close to the minimum in the relative $L^2$ error. Due
to this we say that the estimate of the effective sample size found by
averaging the statistic over each iteration is a good indicator for
the optimal scaling parameter. In general this indicator overestimates
the value of $\neff$ found by using the entire sample.

@@

Both adaptive algorithms are run with initial values of
$\beta=1$. Figure~\ref{fig:MH1 L2} shows that after an initial burn-in
period the APAIS-RW algorithm catches up to the PAIS-RW algorithm, and
by the end of the simulation window is matching its performance. The
ARWMH algorithm does not perform quite as well, this is possibly due
to the fact that the acceptance rate cost function is not particularly
smooth at the optimal value making it difficult to minimise.

@@

Figure~\ref{fig:PAIS_saving} was produced using the method of finding
optimal $\beta$ described in Section~\ref{sec:Implementation P1} (1),
then running 32 repeats at each ensemble size. The convergence rates
are then found by regressing through the data. The graph is still very
noisy but demonstrates that increasing the ensemble size continues to
reduce the number of iterations required in comparison with naively
parallelised MH. The decreasing trend indicates superlinear improvement of PAIS with
respect to ensemble size, in terms of the number of iterations
required, which is a demonstration of our belief that parallelism of
MCMC should give us added value over and above that provided by naive parallelism. This decrease is linked to the increasing effective sample size shown in Figure~\ref{fig:neff-M} (b).

>>>>

In this section we demonstrate convergence of the PAIS algorithm. We
also investigate the convergence properties of PAIS compared with
naively parallelised Metropolis-Hastings samplers. We assess the
performance of the PAIS algorithm using the relative $L^2$ error
defined in \eqref{eqn:L2_error}, as well as the relative error in the
first moment \eqref{eq:34567}.

In the one-dimensional examples which follow, we have optimised the
naively parallelised RWMH algorithm using the optimal acceptance rate
$\hat{\alpha} = 0.5$. This value differs from the theoretical
asymptotic value of 0.234 which applies in higher dimensions, but this
higher acceptance rate is commonly used for one dimensional Gaussian
posteriors~\cite{rosenthal2011optimal}. To find the optimal scaling
parameter we minimise the statistic
\[
	T_{\text{MH}}(\beta) = \left| \frac{N_{\text{acc}}(\beta)}{N_{\text{total}}} - \hat{\alpha} \right|,
\]
where $N_{\text{acc}}(\beta)$ is the number of accepted moves and
$N_{\text{total}}$ is the total number of samples produced. For the
PAIS algorithm, we maximise the effective sample size as
discussed in Section~\ref{sec:ess}.

@@

This first example shows how PAIS handles searching for the mode of a
Gaussian distribution when the initial state is a long way out in the
tails.

@@

Consider the simple case of a linear observation operator $\G(x) = x$,
where the prior on $x$ and the observational noise follow centred
Gaussian distributions. Then, following \eqref{eqn:like}, the Gaussian
posterior has the form
\begin{equation}\label{eqn:Gaussian posterior}
	\text{law}(\mu_D) = \pi(x|D) \propto \exp\left(-\frac{1}{2}\big\|x - D\big\|^2_{\sigma^2} - \frac{1}{2}\big\|x\big\|^2_{\tau^2}\right),
\end{equation}
where $\sigma^2$ and $\tau^2$ are the variances of the
observational noise and prior distributions respectively. In the
numerics which follow, we choose $\tau^2 =0.01$ and $\sigma^2 = 0.01$,
and we observe $x_\text{ref}=4$ noisily such that
\[
D = \mathcal{G}(x) + \eta \sim \mathcal{N}(\mathcal{G}(x_\text{ref}),\sigma^2) = \mathcal{N}(4,0.01).
\]
These values result in a posterior density in which the vast majority
of the density is out in the tails of the
prior distribution. The Kullback-Leibler (KL)
divergence, which gives us a measure of how different the prior and
posterior are, is $D_{KL}(\mu_D || \mu_0) = 4.67$ for this
problem. A KL divergence of zero indicates that two distributions are
identical almost everywhere.

@@

\begin{figure}[htb]
\centering
\subfigure[RWMH algorithm.]{\includegraphics[width=0.45\textwidth]{"figures/G2_rwmho"}}
\subfigure[PAIS-RW algorithm.]{\includegraphics[width=0.45\textwidth]{"figures/G2_paiso"}}
\caption{Finding optimal values of $\beta$ for the example in Section~\ref{sec:problem 1}. The setup is as in Section~\ref{sec:Implementation P1}. Resampling is performed using the ETPF.}
\label{fig:P1 opt beta}
\end{figure}

@@

Figure~\ref{fig:P1 opt beta} (b) shows the effective sample size ratio
compared to the error analysis and the variance of the weights. The
relative $L^2$ error graph is noisy, but it is clear that the maximum
in the effective sample size and the minimum in the variance of the
weights are both close to the minimum in the relative $L^2$ error. Due
to this we say that the estimate of the effective sample size found by
averaging the statistic over each iteration is a good indicator for
the optimal scaling parameter.

@@

Both adaptive algorithms are run with initial values of
$\beta=1$. Figure~\ref{fig:MH1 L2} shows that after an initial burn-in
period the APAIS-RW algorithm catches up to the PAIS-RW algorithm, and
by the end of the simulation window is matching its performance. The
ARWMH algorithm does not perform quite as well, this is possibly due
to the fact that $T_{\text{MH}}(\beta)$ is not smooth at the optimal
value making it difficult to minimise.

@@

Figure~\ref{fig:PAIS_saving} was produced using the method of finding
optimal $\beta$ described in Section~\ref{sec:Implementation P1} (1),
then running 32 repeats at each ensemble size. The convergence rates
are then found by regressing through the data. The graph is still very
noisy but demonstrates that increasing the ensemble size continues to
reduce the number of iterations required in comparison with naively
parallelised MH. The decreasing trend shows superlinear improvement of
PAIS with respect to ensemble size, in terms of the number of
iterations required, which is a demonstration of our belief that
parallelism of MCMC should give us added value over and above that
provided by naive parallelism. This decrease is due to the increasing
effective sample size shown in Figure~\ref{fig:neff-M} (b).
